---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mighty.component

<!-- badges: start -->
[![R-CMD-check](https://github.com/NN-OpenSource/mighty.component/actions/workflows/check_and_co.yaml/badge.svg)](https://github.com/NN-OpenSource/mighty.component/actions/workflows/check_and_co.yaml)
<!-- badges: end -->

mighty.component serve as a repository of generic compute component classes used to produce ADaM scripts in the {mighty} framework.

## Installation

You can install the development version of `mighty.component` from GitHub with:

```{r, eval=FALSE}
pak::pak("NN-OpenSource/mighty.component")
```

## Usage

`mighty.component` provide generic classes to work with standard components, and helper functions to retrieve them.
For now it also holds a small temporary library of components.

See `vignette("mighty-standards")` on how to work with standard components.
