---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mighty.standards

<!-- badges: start -->
[![R-CMD-check](https://github.com/NN-OpenSource/mighty.standards/actions/workflows/check_and_co.yaml/badge.svg)](https://github.com/NN-OpenSource/mighty.standards/actions/workflows/check_and_co.yaml)
<!-- badges: end -->

mighty.standards serve as a repository of generic compute nodes used to produce ADaM scripts in the {mighty} framework.

## Installation

You can install the development version of `mighty.standards` from GitHub with:

```{r, eval=FALSE}
pak::pak("NN-OpenSource/mighty.standards")
```

## Usage

Functions inside `mighty.standards` are not meant to be called directly, 
but instead used as input to the `mighty` framework for automatic generation of ADaM programs.

See `vignette("mighty-standards")` on how to create new standard nodes.
