---
title: "Get started"
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r setup, include = FALSE}
library(mighty.component)
library(testthat)
```

## Overview

mighty.component is the component engine for the
[mighty](https://novonordisk-opensource.github.io/mighty/) framework.
It retrieves, renders, and validates reusable code components used to
build ADaM datasets.

Components are small, self-contained pieces of R code with structured
metadata (dependencies, outputs, type). They come in two formats:

- **Mustache templates** (`.mustache`) -- parameterized templates that
  produce concrete R code when rendered with specific values.
- **R scripts** (`.R`) -- plain R code with no parameters.

This vignette walks through discovering, retrieving, rendering, and
writing components.

## Browse available standards

Standard components ship with the package and cover common ADaM
derivations. Use `list_standards()` to see what is available:

```{r}
list_standards()
```

For more detail, pass `as = "tibble"`:

```{r}
list_standards(as = "tibble")
```

## Retrieve a standard component

`get_standard()` returns a `mighty_component` object -- the unrendered
template with its metadata:
```{r}
ady <- get_standard("ady")
ady
```

The printed output shows the component ID, description, type,
parameters, dependencies, and outputs. You can access each field
directly:

```{r}
ady$title
ady$params
ady$depends
ady$outputs
```

The `$code` field contains the raw template with mustache placeholders:

```{r}
ady$code
```

## Render a component

Rendering replaces mustache placeholders with concrete values and
returns a `mighty_component_rendered` object:

```{r}
rendered <- get_rendered_standard(
  "ady",
  list(domain = "adae", variable = "ASTDY", date = "ASTDT")
)
rendered
```

The rendered `$code` is plain R, ready to execute:

```{r}
rendered$code
```

`get_rendered_component()` is a convenience wrapper that accepts both
standard names and file paths:

```{r}
get_rendered_component(
  "assign",
  list(domain = "adsl", variable = "STUDYID", value = '"XXXX-YY-001"')
)
```

## Write a custom component

### R script format

An R script component (`.R` file) contains fixed derivation logic with
no parameters. The metadata header uses roxygen-style tags, and all
executable code follows the `@code` tag.

```{r, eval = FALSE}
#' @title Age group classification
#' @description Categorizes subjects into age groups based on AGE.
#' @type derivation
#' @depends ADSL AGE
#' @outputs AGEGR1
#' @code
ADSL <- ADSL |>
  dplyr::mutate(
    AGEGR1 = dplyr::case_when(
      AGE < 18 ~ "<18",
      AGE <= 64 ~ "18-64",
      AGE <= 75 ~ "65-75",
      TRUE ~ ">75"
    )
  )
```

Save this as, say, `components/der_agegr1.R` and load it with
`get_component()`:

```{r}
# Create a temporary R script component for demonstration
tmp <- tempfile(fileext = ".R")
writeLines(
  c(
    "#' @title Age group classification",
    "#' @description Categorizes subjects into age groups based on AGE.",
    "#' @type derivation",
    "#' @depends ADSL AGE",
    "#' @outputs AGEGR1",
    "#' @code",
    "ADSL <- ADSL |>",
    '  dplyr::mutate(AGEGR1 = dplyr::case_when(',
    '    AGE < 18 ~ "<18",',
    '    AGE <= 64 ~ "18-64",',
    '    AGE <= 75 ~ "65-75",',
    '    TRUE ~ ">75"',
    "  ))"
  ),
  tmp
)

get_component(tmp)
```

### Mustache template format

A Mustache template (`.mustache` file) adds `@param` tags and uses
`{{placeholders}}` in both the metadata and code sections. This lets the
same logic be reused with different inputs.

```mustache
#' @title Analysis relative day
#' @description
#' Derives the relative day compared to the treatment start date.
#'
#' @param domain Name of the ADaM domain
#' @param variable Name of new variable to create
#' @param date Name of date variable to use
#' @type derivation
#' @depends {{domain}} {{date}}
#' @depends {{domain}} TRTSDT
#' @outputs {{variable}}
#' @code
{{domain}} <- {{domain}} |>
  dplyr::mutate(
    {{variable}} = admiral::compute_duration(
      start_date = TRTSDT,
      end_date = {{date}},
      in_unit = "days",
      out_unit = "days",
      add_one = TRUE
    )
  )
```

### Metadata tags reference

| Tag | Description |
|---|---|
| `@title` | Short title for the component |
| `@description` | Detailed explanation of what the component does |
| `@param` | Template parameter (Mustache templates only) |
| `@type` | Component type (`derivation`, `row`, or `assigned`) |
| `@depends` | Input dependency: `DOMAIN COLUMN` (repeat for each) |
| `@outputs` | Column(s) created or modified (repeat for each) |
| `@code` | Marks the start of executable code |

## Test a component

`get_test_component()` creates a `mighty_component_test` object that
runs component code in an isolated R session with coverage tracking.
The workflow is:

1. Create the test component with `get_test_component()`
2. Assign input data with `$assign()`
3. Evaluate with `$eval()`
4. Retrieve results with `$get()`
5. Check with testthat expectations

Here is an example testing the `ady` standard:

```{r}
test_that("ady derives analysis relative day", {
  # 1. Create the test component
  cmp <- get_test_component(
    "ady",
    params = list(domain = "adae", variable = "ASTDY", date = "ASTDT"),
    check_coverage = FALSE
  )

  # 2. Assign input data
  cmp$assign(
    "adae",
    data.frame(
      ASTDT = as.Date(c("2024-01-10", "2024-01-01", "2023-12-25")),
      TRTSDT = as.Date(c("2024-01-01", "2024-01-01", "2024-01-01"))
    )
  )
  # 3. Evaluate
  cmp$eval()
  # 4. Retrieve results
  result <- cmp$get("adae")

  # 5. Check with testthat expectations
  expect_equal(result$ASTDY, c(10, 1, -7))
})
```

`get_test_component()` registers a deferred `$check_coverage()` call
that verifies every line of the component code was executed. You can
also inspect coverage directly with `$percent_coverage` and
`$line_coverage`.

See `vignette("testing-components")` for a detailed guide on writing
component tests.
