---
title: "Getting Started with mighty.component"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mighty.component}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mighty.component)
```

## What is a mighty component?

Components let you write a data transformation once and reuse it across studies
by swapping variable names at render time. Instead of copying and modifying code
for each new study, you maintain a single template.

A mighty component is a reusable code template for a single, well-defined data
transformation step. Components are commonly used to generate ADaM (Analysis
Dataset Model) programs, but the concept is general: any parameterized R code
snippet that reads a data set, modifies it, and writes it back can be expressed
as a component.

Components combine two ideas:

- **Mustache templating** тАФ placeholders like `{{ domain }}` are filled in at
  render time, so the same logic works across different data sets and variables.
- **Roxygen-like documentation** тАФ tags like `@title`, `@param`, and `@depends`
  describe what the component does, what it needs, and what it produces.

Think of components as reusable building blocks: each one handles a single
derivation or transformation, and you compose several of them to build a
complete program.

## Anatomy of a component template

Below is a minimal component that doubles a column. Every tag is visible at a
glance:
```
#' @title Double a variable
#' @description
#' Creates a new column that is twice the value of an existing column.
#'
#' @param domain `character` Name of the domain (data frame)
#' @param input `character` Name of the existing column to double
#' @param output `character` Name of the new column to create
#' @type derivation
#' @depends {{domain}} {{input}}
#' @outputs {{output}}
#' @code
{{domain}} <- {{domain}} |>
  dplyr::mutate(
    {{output}} = 2 * {{input}}
  )
```

### Tags reference

| Tag | Purpose |
|-----|---------|
| `@title` | One-line title (required) |
| `@description` | Multi-line description (required) |
| `@param name description` | Declares a Mustache placeholder the user must supply at render time |
| `@type` | Semantic type: `derivation`, `predecessor`, `assigned`, or `row` |
| `@depends domain column` | Declares that the code reads `column` from `domain` (repeat for each) |
| `@outputs variable` | Declares a column the code creates (repeat for each) |
| `@code` | Everything below this tag is executable R code |

### Mustache syntax

Inside `@code`, `{{ }}` are **Mustache placeholders** тАФ not R syntax. They are
text-replaced with concrete values before the code is parsed as R.

- **`{{ variable }}`** тАФ replaced with the value supplied at render time.
- **`{{{ variable }}}`** тАФ unescaped rendering, used when the value is literal R
  code (e.g., `{{{ value }}}` in an assignment component to pass `1`, `"text"`,
  or an expression).
- **`{{#list}}...{{/list}}`** тАФ iterates over a vector parameter. Used in
  components that accept multiple values (e.g., the `predecessor` standard
  iterates over `by` and `variable`).

### Conventions

1. The input data set is always called `{{ domain }}`.
2. The code must assign the result back to `{{ domain }}`.
3. Use explicit package namespaces (`dplyr::mutate()`, not `mutate()`).
4. Joins must always specify an explicit `by` argument тАФ this is enforced by
   automatic validation (see [Automatic code validation]).

## Retrieve and inspect a component

List the built-in standard components:

```{r list-standards}
list_standards()
```

Retrieve one by name:

```{r get-ady}
comp <- get_component("ady")
comp
```

Access individual fields through the active bindings:

```{r inspect-fields}
comp$title
comp$type
comp$params
comp$depends
comp$outputs
```

You can also retrieve a component from a local `.mustache` or `.R` file by
passing the file path to `get_component()`.

`get_component()` is the recommended entry point тАФ it handles both built-in
standards and custom files. If you know you only need a built-in standard, you
can also use `get_standard()` and `get_rendered_standard()` directly.

## Render a component

Rendering fills in the Mustache placeholders with concrete values. The
`$render()` method takes parameters as named arguments and returns a
`mighty_component_rendered` object:

```{r render-ady}
rendered <- comp$render(domain = "adae", variable = "ASTDY", date = "ASTDT")
rendered
```

Inspect the rendered code тАФ every `{{ }}` placeholder is now a concrete name:

```{r rendered-code}
rendered$code
```

A convenience function combines retrieval and rendering in one step. Note that
`get_rendered_component()` takes parameters as a named `list`, unlike
`$render()` which takes `...`:

```{r shortcut}
rendered2 <- get_rendered_component(
  "ady",
  list(domain = "adae", variable = "ASTDY", date = "ASTDT")
)
rendered2$code
```

If you omit a required parameter, you get an informative error:

```{r render-error, error=TRUE}
comp$render(domain = "adae")
```

## Evaluate rendered code

Once rendered, call `$eval()` to execute the code in your current environment.
The component code contains an assignment (e.g., `adae <- adae |> ...`), and
`$eval()` evaluates that code in the calling environment via
`eval(envir = parent.frame())`. This means `$eval()` modifies the domain
variable in place тАФ no assignment of the return value is needed.

```{r eval-setup}
adae <- pharmaverseadam::adae |>
  dplyr::select(USUBJID, ASTDT, TRTSDT)

names(adae)
```

The `ASTDY` column does not exist yet. Run the rendered component:

```{r eval-run}
rendered$eval()
names(adae)
head(adae)
```

`$eval()` executes the rendered code in the calling environment by default. You
can pass a different environment via the `envir` argument if needed.

If you want to save the rendered code to a script file instead of evaluating it
interactively, use `$stream(path)` to append the code to an R file.

## Writing a custom component

You can author your own components as `.mustache` files. Here is a realistic
example that derives a ratio of the current value to baseline (`R2BASE`) for a
lab parameter. Save the following template to a `.mustache` file:

```
#' @title Ratio to baseline
#' @description
#' Derives the ratio of the analysis value to the baseline value.
#'
#' @param domain `character` Name of the domain
#' @param variable `character` Name of the new ratio variable
#' @type derivation
#' @depends {{domain}} AVAL
#' @depends {{domain}} BASE
#' @outputs {{variable}}
#' @code
{{domain}} <- {{domain}} |>
  dplyr::mutate(
    {{variable}} = dplyr::if_else(BASE != 0, AVAL / BASE, NA_real_)
  )
```

```{r custom-write}
r2base_file <- tempfile(fileext = ".mustache")
writeLines(c(
  "#' @title Ratio to baseline",
  "#' @description",
  "#' Derives the ratio of the analysis value to the baseline value.",
  "#'",
  "#' @param domain `character` Name of the domain",
  "#' @param variable `character` Name of the new ratio variable",
  "#' @type derivation",
  "#' @depends {{domain}} AVAL",
  "#' @depends {{domain}} BASE",
  "#' @outputs {{variable}}",
  "#' @code",
  "{{domain}} <- {{domain}} |>",
  "  dplyr::mutate(",
  "    {{variable}} = dplyr::if_else(BASE != 0, AVAL / BASE, NA_real_)",
  "  )"
), r2base_file)
```

Load it, render it, and run it:

```{r custom-load}
r2base_comp <- get_component(r2base_file)
r2base_comp
```

```{r custom-render}
r2base_rendered <- r2base_comp$render(
  domain = "adlb",
  variable = "R2BASE"
)
r2base_rendered$code
```

```{r custom-eval}
adlb <- pharmaverseadam::adlb |>
  dplyr::filter(PARAMCD == "ALB") |>
  dplyr::select(USUBJID, PARAMCD, AVISIT, AVAL, BASE)

head(adlb)

r2base_rendered$eval()

adlb |>
  dplyr::select(USUBJID, PARAMCD, AVISIT, AVAL, BASE, R2BASE) |>
  head()
```

## Automatic code validation

When a component is rendered, the generated code is automatically validated.
The package currently checks for **implicit joins** тАФ any `dplyr::left_join()`,
`dplyr::inner_join()`, or similar call without an explicit `by` argument triggers
an error. This prevents a common source of bugs in clinical programming where
join columns change between studies.

Here is a component that fails validation:

```{r validation-fail, error=TRUE}
bad_template <- c(
  "#' @title Bad join example",
  "#' @description Implicit join that will fail validation.",
  "#'",
  "#' @param domain `character` domain name",
  "#' @type predecessor",
  "#' @depends {{domain}} USUBJID",
  "#' @outputs NEWCOL",
  "#' @code",
  "{{domain}} <- {{domain}} |>",
  "  dplyr::left_join(other_data)"
)

bad_file <- tempfile(fileext = ".mustache")
writeLines(bad_template, bad_file)
get_rendered_component(bad_file, list(domain = "adae"))
```

The fix is to specify the join key explicitly:

```{r validation-pass}
good_template <- c(
  "#' @title Good join example",
  "#' @description Explicit join that passes validation.",
  "#'",
  "#' @param domain `character` domain name",
  "#' @type predecessor",
  "#' @depends {{domain}} USUBJID",
  "#' @outputs NEWCOL",
  "#' @code",
  "{{domain}} <- {{domain}} |>",
  "  dplyr::left_join(other_data, by = dplyr::join_by(USUBJID))"
)

good_file <- tempfile(fileext = ".mustache")
writeLines(good_template, good_file)
good_comp <- get_rendered_component(good_file, list(domain = "adae"))
good_comp$code
```

## Testing components

`get_test_component()` creates a component that runs in an **isolated R session**
with automatic code coverage tracking. This is useful both for interactive
exploration and for formal unit tests with `testthat`.

We set `check_coverage = FALSE` here because this code runs inside a vignette,
not inside a `test_that()` block. The default (`TRUE`) uses `withr::defer()` to
automatically verify coverage when a test finishes тАФ use that default in your
actual tests.

```{r test-create}
test <- get_test_component(
  component = "ady",
  params = list(domain = "adae", variable = "ASTDY", date = "ASTDT"),
  check_coverage = FALSE
)
test
```

Assign input data into the isolated session:

```{r test-assign}
adae_input <- pharmaverseadam::adae |>
  dplyr::select(USUBJID, ASTDT, TRTSDT)

test$assign("adae", adae_input)
test$ls()
```

Execute the component and retrieve the result:

```{r test-eval}
test$eval()
result <- test$get("adae")
head(result)
```

Check coverage тАФ every line of the component code should have been executed:

```{r test-coverage}
test$percent_coverage
test$line_coverage
```

When `check_coverage = TRUE` (the default), coverage is verified automatically
when the test object goes out of scope using `withr::defer()`. If any line was
not executed, an error is raised. This integrates naturally with `testthat` test
files: create the test component inside a `test_that()` block, assign data,
evaluate, and assert on the results тАФ coverage checking happens automatically
when the test finishes.

## Chaining multiple components

Components are designed to be composed. Each one takes a data frame, adds or
modifies columns, and passes it along. Here is an example of a typical
SDTM-to-ADaM workflow: start with the SDTM AE domain, merge subject-level
treatment dates from ADSL, then derive ADaM analysis variables step by step.

```{r chain-setup}
adae <- pharmaversesdtm::ae |>
  dplyr::select(USUBJID, AESEQ, AETERM, AESTDTC, AEENDTC)

adsl <- pharmaverseadam::adsl |>
  dplyr::select(USUBJID, TRTSDT, TRTEDT)

adae <- adae |>
  dplyr::left_join(adsl, by = dplyr::join_by(USUBJID))

names(adae)
```

Derive analysis start date, end date, and relative day:

```{r chain-derive}
get_rendered_component(
  "astdt",
  list(domain = "adae", dtc = "AESTDTC")
)$eval()

get_rendered_component(
  "aendt",
  list(domain = "adae", dtc = "AEENDTC")
)$eval()

get_rendered_component(
  "ady",
  list(domain = "adae", variable = "ASTDY", date = "ASTDT")
)$eval()
```

Inspect the result:

```{r chain-result}
adae |>
  dplyr::select(
    USUBJID, AESTDTC, ASTDT, ASTDTF, AEENDTC, AENDT, AENDTF, ASTDY
  ) |>
  head()
```

Each `get_rendered_component()$eval()` call is one composable step. In the full
mighty framework these steps are orchestrated automatically, but you can also
use them interactively to build programs step by step.

## Summary

The mighty component workflow:

1. **Write** a component template with Mustache placeholders and documentation
   tags.
2. **Retrieve** it with `get_component()`.
3. **Render** it with `$render()` to fill in the placeholders.
4. **Evaluate** it with `$eval()` to run the code against your data.
5. **Test** it with `get_test_component()` for isolated execution and coverage
   tracking.

Use `list_standards(as = "tibble")` for a quick overview of all built-in
standard components. See the [package reference
site](https://novonordisk-opensource.github.io/mighty.component) for full API
documentation.
