---
title: "Getting Started with mighty.component"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mighty.component}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mighty.component)
```

## What is a component?

A component is a reusable code template for one data transformation step. Its
code takes an input data set and returns it with new or changed columns or
rows. Mustache placeholders, such as `{{{domain}}}`, are filled in when the
component is rendered, so the same template works across studies. Roxygen-like
tags, such as `@title`, `@param` and `@depends`, document what the component
does, needs and creates.

Components are commonly used to build ADaM (Analysis Data Model) programs. Each
component handles one derivation, and several components make up a program. In
the mighty ecosystem, mighty.metadata provides study configuration (via
[`mighty_study()`](https://novonordisk-opensource.github.io/mighty.metadata/reference/mighty_study.html)
and `_study.yml`) that controls which components are rendered and with which
parameters.

## Anatomy of a component template

Below is a minimal component that creates a column as two times an existing
column:

```r
#' @title Double a column
#' @description
#' Creates a new column as two times an existing column.
#'
#' @param domain `character` Name of the domain
#' @param output `character` Name of the new column
#' @param input `character` Name of the existing column
#' @type column
#' @origin Derived
#' @method Two times the input column
#' @depends {{{domain}}} {{{input}}}
#' @outputs {{{output}}}
#' @code
{{{domain}}} <- {{{domain}}} |>
  dplyr::mutate(
    {{{output}}} = 2 * {{{input}}}
  )
```

The tags at the top describe the component. `@title`, `@description`,
`@type`, `@origin`, `@method` and `@code` are required. `@param` declares
a placeholder, `@depends` a column the code reads, and `@outputs` a column it
creates. All lines below `@code` are the R code.

Templates use [Mustache](https://mustache.github.io/mustache.5.html)
placeholders. `{{{name}}}` inserts the value of parameter `name`. Always use
triple braces without inner spaces: double braces HTML-escape the value, and
spaces break `@depends`. `{{#name}}...{{/name}}` repeats the enclosed text for
each element of the vector `name`, and `{{{.}}}` inserts the current element.

See `?mighty_component` for the full tag reference, component types,
conventions and validation rules.

## Retrieve and inspect a component

List the example components shipped with the package:

```{r list-components}
path <- system.file("examples", package = "mighty.component")
list_components(path)
```

Retrieve one by file path:

```{r get-ady}
ady <- get_component(
  system.file("examples", "ady.mustache", package = "mighty.component")
)
ady
```

Access individual fields:

```{r inspect-fields}
ady$id
ady$title
ady$type
ady$params
ady$depends
ady$outputs
ady$origin
ady$method
```

## Render a component

Rendering fills in the placeholders with values. `$render()` takes one named
argument per `@param` tag and returns a `mighty_component_rendered` object.
Every placeholder is now a name:

```{r render-ady}
ady_rendered <- ady$render(domain = "ADAE", variable = "ASTDY", date = "ASTDT")
ady_rendered
```

The rendered code is validated: it must be valid R, and joins must specify
`by` (see [Automatic code validation]). Columns are not checked. Use
`get_test_component()` (see [Testing components]) to run a component on real
data.

`get_rendered_component()` retrieves and renders in one step. It takes the
parameters as a named `list`:

```{r shortcut}
get_rendered_component(
  system.file("examples", "ady.mustache", package = "mighty.component"),
  list(domain = "ADAE", variable = "ASTDY", date = "ASTDT")
)
```

A missing parameter raises an error:

```{r render-error, error=TRUE}
ady$render(domain = "ADAE")
```

## Evaluate rendered code

`$eval()` runs the code. By default it runs in the calling environment. Use the
`envir` argument to choose another environment. The code assigns the result
back to the domain (e.g., `ADAE <- ADAE |> ...`), so `ADAE` is updated without
assigning the return value.

```{r eval-setup}
ADAE <- pharmaverseadam::adae |>
  dplyr::select(USUBJID, ASTDT, TRTSDT)

names(ADAE)
```

`ASTDY` does not exist yet. Run the rendered component:

```{r eval-run}
ady_rendered$eval()
names(ADAE)
head(ADAE)
```

`$stream()` appends the code to an R script instead:

```{r stream-example}
script_file <- tempfile(fileext = ".R")
ady_rendered$stream(script_file)
readLines(script_file)
```

## Writing a custom component

Write your own components as `.mustache` files. This component derives the
ratio of the analysis value to baseline:

```r
#' @title Ratio to baseline
#' @description
#' Derives the ratio of the analysis value to the baseline value.
#'
#' @param domain `character` Name of the domain
#' @param variable `character` Name of the new ratio variable
#' @type column
#' @origin Derived
#' @method Ratio of AVAL to BASE
#' @depends {{{domain}}} AVAL
#' @depends {{{domain}}} BASE
#' @outputs {{{variable}}}
#' @code
{{{domain}}} <- {{{domain}}} |>
  dplyr::mutate(
    {{{variable}}} = dplyr::if_else(BASE != 0, AVAL / BASE, NA_real_)
  )
```

```{r custom-write, include = FALSE}
r2base_file <- file.path(tempdir(), "r2base.mustache")
writeLines(c(
  "#' @title Ratio to baseline",
  "#' @description",
  "#' Derives the ratio of the analysis value to the baseline value.",
  "#'",
  "#' @param domain `character` Name of the domain",
  "#' @param variable `character` Name of the new ratio variable",
  "#' @type column",
  "#' @origin Derived",
  "#' @method Ratio of AVAL to BASE",
  "#' @depends {{{domain}}} AVAL",
  "#' @depends {{{domain}}} BASE",
  "#' @outputs {{{variable}}}",
  "#' @code",
  "{{{domain}}} <- {{{domain}}} |>",
  "  dplyr::mutate(",
  "    {{{variable}}} = dplyr::if_else(BASE != 0, AVAL / BASE, NA_real_)",
  "  )"
), r2base_file)
```

Save the template as `r2base.mustache`. Then load, render and run it:

```{r custom-load}
r2base <- get_component(r2base_file)
r2base
```

```{r custom-render}
r2base_rendered <- r2base$render(
  domain = "ADLB",
  variable = "R2BASE"
)
r2base_rendered$code
```

```{r custom-eval}
ADLB <- pharmaverseadam::adlb |>
  dplyr::filter(PARAMCD == "ALB") |>
  dplyr::select(USUBJID, PARAMCD, AVISIT, AVAL, BASE)

head(ADLB)

r2base_rendered$eval()

ADLB |>
  dplyr::select(USUBJID, PARAMCD, AVISIT, AVAL, BASE, R2BASE) |>
  head()
```

## Automatic code validation

Rendered code is validated. It must be valid R, and joins from dplyr, tidylog
and dbplyr must specify `by`. Implicit joins are a common source of bugs when
key columns differ between studies.

This component fails validation:

```r
#' @title Bad join example
#' @description Implicit join that will fail validation.
#'
#' @param domain `character` Name of the domain
#' @type row
#' @origin Derived
#' @method Implicit join on other_data
#' @depends {{{domain}}} USUBJID
#' @outputs NEWCOL
#' @code
{{{domain}}} <- {{{domain}}} |>
  dplyr::left_join(other_data)
```

```{r validation-fail-setup, include = FALSE}
bad_template <- c(
  "#' @title Bad join example",
  "#' @description Implicit join that will fail validation.",
  "#'",
  "#' @param domain `character` Name of the domain",
  "#' @type row",
  "#' @origin Derived",
  "#' @method Implicit join on other_data",
  "#' @depends {{{domain}}} USUBJID",
  "#' @outputs NEWCOL",
  "#' @code",
  "{{{domain}}} <- {{{domain}}} |>",
  "  dplyr::left_join(other_data)"
)

bad_file <- file.path(tempdir(), "bad_join.mustache")
writeLines(bad_template, bad_file)
```

```{r validation-fail, error=TRUE}
get_rendered_component(bad_file, list(domain = "ADAE"))
```

Specify the join key with `by` to fix it:

```r
#' @title Good join example
#' @description Explicit join that passes validation.
#'
#' @param domain `character` Name of the domain
#' @type row
#' @origin Derived
#' @method Explicit join on other_data by USUBJID
#' @depends {{{domain}}} USUBJID
#' @outputs NEWCOL
#' @code
{{{domain}}} <- {{{domain}}} |>
  dplyr::left_join(other_data, by = dplyr::join_by(USUBJID))
```

```{r validation-pass-setup, include = FALSE}
good_template <- c(
  "#' @title Good join example",
  "#' @description Explicit join that passes validation.",
  "#'",
  "#' @param domain `character` Name of the domain",
  "#' @type row",
  "#' @origin Derived",
  "#' @method Explicit join on other_data by USUBJID",
  "#' @depends {{{domain}}} USUBJID",
  "#' @outputs NEWCOL",
  "#' @code",
  "{{{domain}}} <- {{{domain}}} |>",
  "  dplyr::left_join(other_data, by = dplyr::join_by(USUBJID))"
)

good_file <- file.path(tempdir(), "good_join.mustache")
writeLines(good_template, good_file)
```

```{r validation-pass}
get_rendered_component(good_file, list(domain = "ADAE"))$code
```

## Testing components

`get_test_component()` creates a component that runs in a separate R session
and tracks code coverage. Use it for interactive checks and for unit tests with
testthat.

By default (`check_coverage = TRUE`), coverage is checked when the calling
environment ends, e.g. the `test_that()` block. An error is raised if any line
has not run. This vignette is not a test, so it sets `check_coverage = FALSE`.

```{r test-create}
ady_path <- system.file(
  "examples", "ady.mustache",
  package = "mighty.component"
)
ady_test <- get_test_component(
  component = ady_path,
  params = list(domain = "ADAE", variable = "ASTDY", date = "ASTDT"),
  check_coverage = FALSE # keep the default TRUE in tests
)
ady_test
```

Assign input data in the test session:

```{r test-assign}
ADAE_input <- pharmaverseadam::adae |>
  dplyr::select(USUBJID, ASTDT, TRTSDT)

ady_test$assign("ADAE", ADAE_input)
ady_test$ls()
```

Run the component and get the result:

```{r test-eval}
ady_test$eval()
ady_test$get("ADAE") |> head()
```

Every line of the code has now run:

```{r test-coverage}
# Print method
ady_test
# Percent coverage
ady_test$percent_coverage
# Line coverage as a data.frame
ady_test$line_coverage
```

In a testthat file, create the test component inside `test_that()`, assign
data, evaluate and test the results. Coverage is checked when the test ends.
See `?mighty_component_test` for the workflow.
